---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  out.width = "100%"
)

set.seed(123L)
```

# incubate

<!-- badges: start -->
<!-- badges: end -->

Parametric time-to-event analysis where groups show an incubation period with different hazard.


## Example

With `incubate`, you can statistically compare the survival experience of two groups with respect to delay and other parameters.

```{r example, out.width='60%'}
library(incubate)

# simulate data from exponential distribution with delay
x <- rexp_delayed(n = 13, delay = 1.1)
y <- rexp_delayed(n = 11, delay = 1.6)

# fit delay model 
fm <- delay_model(x, y)

plot(fm)

# confidence interval for delay-parameters
confint(fm, parm = c('delay.x', 'delay.y'))

# test on difference in delay, using only R=50 bootstrap draws
delay_test <- test_delay_diff(x, y, R = 50)
plot(delay_test)
```

## Parallel computation

To switch on parallel computation, e.g. for bootstrap parameter tests or power simulations, simply set up a suitable computation plan via the Future-API.
For instance, do the following to activate four R-sessions in the background of your local computer for computer-intensive tasks in `incubate`:

```{r future_setup, echo=TRUE, eval=FALSE}
library(future)
plan(multisession, workers = 4)
```

That's it. You do *not* have to change any function calls. `incubate` is `future`-aware.
Consult the [`future`-package on CRAN](https://cran.r-project.org/web/packages/future/) for more information about futures and about supported computation plans in R.

When you are done with the heavy computing, it is best practice to release the parallel connections:  
```{r future_teardown, echo=-1, eval=FALSE}
library(future)
plan(sequential)
```



## Installation

The `incubate` package is hosted publicly at [Gitlab](https://gitlab.com/imb-dev/incubate).
To install its latest version use from within an R-session:

``` r
remotes::install_gitlab("imb-dev/incubate")
```

To install a specific version, add the version tag after the name, separated by a `@`, e.g. to install `incubate` in version `v0.0.3` use

``` r
remotes::install_gitlab("imb-dev/incubate@v0.0.3")
```

