---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

set.seed(1234L)
```

# incubate

<!-- badges: start -->
<!-- badges: end -->

Parametric time-to-event analysis where groups show an incubation period where hazard is different.


## Installation

You can install the released version of incubate from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("incubate")
```

## Example

With incubate, you can statistically compare the survival experience of two groups with respect to delay and other parameters.

```{r example}
library(incubate)

# simulate data from exponential distribution with delay
x <- rexp_delayed(n = 13, delay = 1.1)
y <- rexp_delayed(n = 11, delay = 1.6)

# fit delay model 
fm <- delay_model(x, y)

plot(fm)

delay_test <- test_delay_diff(x, y, R = 100)
plot(delay_test)
```

## Installation

You can install the `sscn` package from [Gitlab](https://gitlab.com/imb-dev/incubate) with:

``` r
remotes::install_gitlab("imb-dev/incubate")
```

To install a specific version, add the version tag after the name, separated by a `@`, e.g. to install `incubate` in version `v0.0.1` use

``` r
remotes::install_gitlab("imb-dev/incubate@v0.0.1")
```

