---
title: "Distributions with Delay"
author: matthias.kuhn@tu-dresden.de
date: 2024-07-25
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributions with Delay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library("dplyr")
library("ggplot2")
library("patchwork")
library("cowplot")
theme_set(cowplot::theme_cowplot())
```

```{r setup}
library("incubate")
```

The delayed exponential distribution and the delayed Weibull distribution
are the foundational data models in the `incubate`-package.

## Parameter transformation
The model fitting uses numeric optimization routines.
We use parameter transformation to improve this process,
in particular through suitable parameter transformation we can avoid constraints on the bounds of the parameters.

For instance, the initial delay parameter $α_1$ is by definition bounded upwards by the first observation $x_{(1)}$
and conventionally we assume $α_1 ≥ 0$ as well, in particular when we model a survival time. This 2^nd^ assumption is not strictly necessary, however, as a model with negative delay might also describe well a survival data.

To get rid of the (upper) bounds, an idea is to use the 1^st^ observation as a normalization for $α_1$ and find a transformation to **whole ℝ**:

For instance, assuming only an upper bound ($α_1 ≤ x_{(1)}$, also negative values allowed) we get a transformation involving the logarithm:
\[
\tilde{α}_1 := \log(\frac{x_{(1)}-α_1}{x_{(1)}}) = \log(1 - \frac{α_1}{x_{(1)}})
\]

Alternatively, if we want to enforce also the lower bound for $α_1$
we can apply a logit-transformation to rescale the delay
from $[0,x_{(1)}]$ to whole ℝ:
\[
\tilde{α}_1 := \operatorname{logit}(\frac{α_1}{x_{(1)}}) = -\log(1 + \frac{x_{(1)} - 2α_1}{α_1})
\]

The back-transformation is $α_1 = \frac{x_{(1)}}{1+\exp(-\tilde{α}_1)}$.
We visualize this logit-transformation:
```{r plot_logit_transf, echo=FALSE, fig.align='center', fig.width=8, fig.height=3}
x1 <- 7

ggplot() + xlim(0, 7) +
  geom_function(fun = \(alpha1) -log1p((x1 - 2 * alpha1)/alpha1), n = 1001) +
  labs(x = expression(alpha), y = expression(tilde(alpha))) |
  
  ggplot() + xlim(-9, 9) +
  geom_function(fun = \(alpha1_tr) x1 / (expm1(-alpha1_tr)+2), n = 1001) + 
  labs(x = expression(tilde(alpha)), y = expression(alpha),
       caption = bquote("transformation for data with " * x[(1)] == .(x1)))
```



## Random right censoring
The functions for random generation of data `rexp_delayed()` and `rweib_delayed()`
have an option to produce data under a random right censoring scheme.
Currently, only random right censoring under a uniform censoring distribution is supported.
The censoring process $C$ is independent of the event process $X$.
You provide the desired expected proportion of right censoring π via parameter `cens=`.
This information determines the upper limit `Z` for the censoring process `U(0,Z)`.
The expected proportion of censorings is $P(C<X)$. Due to the independence we get
\[
P(C<X) = ∫_0^∞ ∫_0^x f_{X,C}(x,c)dc dx = ∫_0^∞ ∫_0^x f_X(x) f_C(c) dc dx = ∫_0^∞ F_C(x) f_X(x)dx
\]

For a uniform censoring process with lower bound 0 and upper bound $Z$
we have $F_C(x) = \frac{x-0}{Z-0} = \frac{x}{Z}$.

### Exponential
Substituting the density $f_X(x) = 𝟙_{x≥α} · λ \exp(-λ·(x-α))$ and using integration by parts
we can solve for $Z$:
\[
π = P(C<X) = \frac{α + λ^{-1}}{Z} \qquad ⇔ \qquad Z = \frac{α + λ^{-1}}{π}
\]

As an example, to draw a random sample from a delayed exponential that on average has 30% right censorings,
we can use the following call:
```{r}
#| label: rcens_exp_exmpl

(x <- rexp_delayed(n = 20, delay1 = 3, rate1 = .2, cens = .3))
class(x)
sum(x[,2] == 0)/length(x)
```


### Weibull
The density of delayed Weibull is
$f_X(x) = 𝟙_{x≥α} · \frac{k}{σ} · (\frac{x-α}{σ})^{k-1} \exp(-(\frac{x-α}{σ})^k)$
with shape parameter $k$ and scale parameter $σ$.
Then, using integration by parts and using 
$∫_0^∞ \exp(-a · x^b) dx = \frac{1}{b} a^{-1/b} · Γ(\frac{1}{b})$
for $a=1$ and $b=k$, the shape parameter,
we get
\[
π = P(C<X) = \frac{α + \frac{σ}{k}·Γ(\frac{1}{k})}{Z} \qquad ⇔ \qquad Z = \frac{α + \frac{σ}{k}·Γ(\frac{1}{k})}{π}
\]
