---
title: "Distributions with Delay"
output: rmarkdown::html_vignette
date: 2023-09-13
vignette: >
  %\VignetteIndexEntry{Distributions with Delay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library("incubate")
```

The delayed exponential distribution and the delayed Weibull distribution
are the foundational data models in the `incubate`-package.


## Random right censoring
The functions for random generation of data `rexp_delayed()` and `rweib_delayed()`
have an option to produce data under a random right censoring scheme.
Currently, only random right censoring under a uniform censoring distribution is supported.
The censoring process $C$ is independent of the event process $X$.
You provide the desired expected proportion of right censoring π via parameter `cens=`.
This information determines the upper limit `Z` for the censoring process `U(0,Z)`.
The expected proportion of censorings is $P(C<X)$. Due to the independence we get
\[
P(C<X) = ∫_0^∞ ∫_0^x f_{X,C}(x,c)dc dx = ∫_0^∞ ∫_0^x f_X(x) f_C(c) dc dx = ∫_0^∞ F_C(x) f_X(x)dx
\]

For a uniform censoring process with lower bound 0 and upper bound $Z$
we have $F_C(x) = \frac{x-0}{Z-0} = \frac{x}{Z}$.

### Exponential
Substituting the density $f_X(x) = 𝟙_{x≥α} · λ \exp(-λ·(x-α))$ and using integration by parts
we can solve for $Z$:
\[
π = P(C<X) = \frac{α + λ^{-1}}{Z} \qquad ⇔ \qquad Z = \frac{α + λ^{-1}}{π}
\]

As an example, to draw a random sample from a delayed exponential that on average has 30% right censorings,
we can use the following call:
```{r}
#| label: rcens_exp_exmpl

(x <- rexp_delayed(n = 20, delay1 = 3, rate1 = .2, cens = .3))
class(x)
sum(x[,2] == 0)/length(x)
```


### Weibull
The density of delayed Weibull is
$f_X(x) = 𝟙_{x≥α} · \frac{k}{σ} · (\frac{x-α}{σ})^{k-1} \exp(-(\frac{x-α}{σ})^k)$
with shape parameter $k$ and scale parameter $σ$.
Then, using integration by parts and using 
$∫_0^∞ \exp(-a · x^b) dx = \frac{1}{b} a^{-1/b} · Γ(\frac{1}{b})$
for $a=1$ and $b=k$, the shape parameter,
we get
\[
π = P(C<X) = \frac{α + \frac{σ}{k}·Γ(\frac{1}{k})}{Z} \qquad ⇔ \qquad Z = \frac{α + \frac{σ}{k}·Γ(\frac{1}{k})}{π}
\]
